Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.Rmd
highlights

EpiCompare emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis. In this section we will highlight weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take. 
To accomplish this goal we will first demonstrate the inability of time-dependent tools to well quantify a classic epidemic parameter, the reproduction number $R_0$. 
Then we will motivate new time-invariant approaches for more complex situations where $R_0$ doesn’t capture the complexities of outbreaks. This will lead us to our final subsection that defines ways to view epidemics in a time-invariant lense and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.

@@

Epidemiological research is often interested in understanding intrinsic properties of the epidemic, independent from when the epidemic occurred or how frequently data was collected. One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$.  This intrinsic property\footnote{added to shorten this long sentence and connect that it is an intrinsic property} is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}. \citet{Gallagher2020} showed that the estimation of $R_0$  can be sensitive to time-dependent parameters like estimation of the beginning and end of an epidemic. This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable, intrinsic properties of an epidemic.

In many situations, epidemiologists want a much deeper understand of an epidemic than just the reproduction number. One common time-dependent tool \shannon{used to better understand epidemics} is a series of time series line plots that track the proportion of the population in a given state (e.g. infected) at a given time. Figure \ref{fig:different-scales-standard} visualizes two different simulated epidemics with population states (S)usceptible, (I)nfectious and (R)ecovered. These simulations have been generated under a discrete approximation of  \citet{Kermack1927}’s SIR model. The Kermack and McKendrick model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,

@@

Even though Figure \ref{fig:different-scales-standard}’s visual analysis might be able to present more complex properties of the epidemic, one might wish to understand how these two simulated epidemics’ $R_0$ values compare. Even though these two simulated epidemics appear different in these figures (including having different infection peaks), there is no real intuitive way to compare the epidemics' $R_0$ values. In fact, both of these simulated epidemics started with the same population (1000 total people with 10 infected) and have the same $R_0$ value, just with slightly different infection and recovery rates ($\beta_1, \gamma_1 = 0.8, 0.4$, and $\beta_2, \gamma_2 = .64, .32$). Even for simple generative models, this time-dependent visualization cannot be used to compare a very simple but powerful numerical summary, $R_0$.

@@

The faceted time series plot, like that seen in Figure \ref{fig:different-scales-standard},  not only fails to allow the practitioner to compare simulated epidemics’ $R_0$ values but also presented epidemics’ trajectory data so that only the marginal information for each state can be examined at a given time.  A more holistic approach to visualizing the overall path of our simulated epidemics is to examine how each epidemic traverses the three-dimensional $(S(t),I(t),R(t))$ space. Figure 3’s left plot does just that, presenting the trajectories of the simulated epidemics in this three-dimensional space. For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size) we can visual these point in a a two-dimensional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}’s right plot. In this both of Figure  \ref{fig:different-scales-tern}’s plots we observe that the two epidemics are on the same trajectory in this three-dimensional space. In this simple generative example setting, being on the same trajectory indicates that \ben{the} two epidemics have the same $R_0$ value. This can be proven mathematically.

@@

\sout{When the goal is to predict the future impact / trajectory of a epidemic,} In addition to being used to better understand completed epidemics, time-invariant tools can be used to aid prediction of future epidemics by providing a state-space regions in which we expect the true epidemic to traverse. In settings where the epidemic only generates a single outbreak, these regions can be very telling if simulation models capture the epidemic’s structure. In \pkg{EpiCompare} we create geometric prediction regions around all but the $\alpha$ proportion of most extreme simulated trajectories.  These geometric regions can also be used to compared simulation models that have different time scales, parameters and even different statistical philosophies, through set different distances. Although visualization is the easiest when the epidemic has three states, prediction regions can be useful to assess and compare simulation models where the epidemics have multiple states.

>>>>

\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis. In this section we highlight some weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take. 
To accomplish this goal we  first demonstrate the inability of time-dependent tools to \sout{well quantify}\shannon{simply capture} a classic epidemic parameter, the reproduction number $R_0$. 
Then we motivate new time-invariant approaches for more complex situations where $R_0$ does not capture the complexities of outbreaks. This leads to the final subsection that defines ways to view epidemics in a time-invariant lens and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.

@@

Epidemiological research is \shannon{often interested}\footnote{\shannon{I used often intsead of "a lot of epidemiological research" which sounds too colloquial}} in understanding intrinsic properties of the epidemic, independent from when the epidemic occurred or how frequently data was collected. One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$. \shannon{} This intrinsic property}\footnote{\shannon{added to shorten this long sentence and connect that it is an intrinsic property}} is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}. \citet{Gallagher2020} showed that the estimation of $R_0$  can be sensitive to time-dependent parameters like estimation of the beginning and end of an epidemic. This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable, \shannon{intrinsic} properties of an epidemic.

In many situations, epidemiologists want a much deeper understand of an epidemic than just the reproduction number. One common time-dependent tool \shannon{used to better understand epidemics} is a series of time series line plots that track the proportion of the population in a given state (e.g. infected) at a given time. Figure \ref{fig:different-scales-standard} visualizes two different simulated epidemics with population states (S)usceptible, (I)nfectious and (R)ecovered. These simulations have been generated under a discrete approximation of  \citet{Kermack1927}’s SIR model. j\shannon{The Kermack and McKendrick} model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,

@@

Even though Figure \ref{fig:different-scales-standard}’s visual analysis might be able to present more complex properties of the epidemic, one might wish to understand how these two simulated epidemics’ $R_0$ values compare. Even though these two simulated epidemics appear different in these figures (including having different infection peaks), there is no real intuitive way to compare the epidemics' $R_0$ values. In fact, both of these simulated epidemics started with the same population (1000 total people with 10 infected) and have the same $R_0$ value, just with slightly different infection and recovery rates ($\beta_1, \gamma_1 = 0.8, 0.4$, and $\beta_2, \gamma_2 = .64, .32$). Even for simple generative models, this time-dependent visualization cannot be used to compare a very simple but powerful numerical summary, \shannon{$R_0$}.

@@

The faceted time series plot, like that seen in Figure \ref{fig:different-scales-standard},  not only fails to allow the practitioner to compare simulated epidemics’ $R_0$ values but also presented epidemics’ trajectory data so that only the marginal information for each state can be examined at a given time.  A more holistic approach to visualizing the overall path of our simulated epidemics is to examine \shannon{how} each epidemic traverses the three-dimensional $(S(t),I(t),R(t))$ space. Figure 3’s left plot does just that, presenting the trajectories of the simulated epidemics in this three-dimensional space. For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size) we can visual these point in a a two-dimensional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}’s right plot. In this both of Figure  \ref{fig:different-scales-tern}’s plots we observe that the two epidemics are on the same trajectory in this three-dimensional space. In this simple generative example setting, \shannon{being on the same trajectory} indicates that \ben{the} two epidemics have the same $R_0$ value. This can be proven mathematically.

@@

\sout{When the goal is to predict the future impact / trajectory of a epidemic,} \shannon{In addition to being used to better understand completed epidemics,} time-invariant tools can be used to aid prediction of future epidemics by providing a state-space regions in which we expect the true epidemic to traverse. In settings where the epidemic only generates a single outbreak, these regions can be very telling if simulation models capture the epidemic’s structure. In \pkg{EpiCompare} we create geometric prediction regions around all but the $\alpha$ proportion of most extreme simulated trajectories.  These geometric regions can also be used to compared simulation models that have different time scales, parameters and even different statistical philosophies, through set different distances. Although visualization is the easiest when the epidemic has three states, prediction regions can be useful to assess and compare simulation models where the epidemics have multiple states.

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.tex
highlights

EpiCompare emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way - approaches that remove some or all of the
impact of start/end times and recording time scales when preforming the
analysis. In this section we will highlight weaknesses of
time-\textit{de}pendent analysis and define the mathematical
underpinning of the time-invariant approach we take. To accomplish this
goal we will first demonstrate the inability of time-dependent tools to
well quantify a classic epidemic parameter, the reproduction number
\(R_0\). Then we will motivate new time-invariant approaches for more
complex situations where \(R_0\) doesn't capture the complexities of
outbreaks. This will lead us to our final subsection that defines ways
to view epidemics in a time-invariant lense and discusses natural
properties of these representations that allow for clearer ways to
compare models and epidemics.

@@

Epidemiological research is often interested in understanding intrinsic
properties of the epidemic, independent from when the epidemic occurred
or how frequently data was collected. One of the most famous numerical
summaries of an epidemic is the reproduction number - \(R_0\). This
intrinsic
property\footnote{added to shorten this long sentence and connect that it is an intrinsic property}
is a time-invariant value that is defined as the expected number of
infections caused by a single infect or who is added to a completely
susceptible population \citep{anderson1992}. \citet{Gallagher2020}
showed that the estimation of \(R_0\) can be sensitive to time-dependent
parameters like estimation of the beginning and end of an epidemic. This
time-dependent sensitivity in estimating \(R_0\) reflects a general
problem in epidemiology surrounding how to transform time-dependent
information into insights into desirable, intrinsic properties of an
epidemic.

In many situations, epidemiologists want a much deeper understand of an
epidemic than just the reproduction number. One common time-dependent
tool \textcolor{orange}{used to better understand epidemics} is a series
of time series line plots that track the proportion of the population in
a given state (e.g.~infected) at a given time. Figure
\ref{fig:different-scales-standard} visualizes two different simulated
epidemics with population states (S)usceptible, (I)nfectious and
(R)ecovered. These simulations have been generated under a discrete
approximation of \citet{Kermack1927}'s SIR model. The Kermack and
McKendrick model captures the transitions from one state to the next as
a system of ordinary differential equations, where \(N\) is the total
number of individuals, \(\beta\) is the rate of infection, and
\(\gamma\) is the rate of recovery,

@@

Even though Figure \ref{fig:different-scales-standard}'s visual analysis
might be able to present more complex properties of the epidemic, one
might wish to understand how these two simulated epidemics' \(R_0\)
values compare. Even though these two simulated epidemics appear
different in these figures (including having different infection peaks),
there is no real intuitive way to compare the epidemics' \(R_0\) values.
In fact, both of these simulated epidemics started with the same
population (1000 total people with 10 infected) and have the same
\(R_0\) value, just with slightly different infection and recovery rates
(\(\beta_1, \gamma_1 = 0.8, 0.4\), and
\(\beta_2, \gamma_2 = .64, .32\)). Even for simple generative models,
this time-dependent visualization cannot be used to compare a very
simple but powerful numerical summary, \(R_0\).

@@

The faceted time series plot, like that seen in Figure
\ref{fig:different-scales-standard}, not only fails to allow the
practitioner to compare simulated epidemics' \(R_0\) values but also
presented epidemics' trajectory data so that only the marginal
information for each state can be examined at a given time. A more
holistic approach to visualizing the overall path of our simulated
epidemics is to examine how each epidemic traverses the
three-dimensional \((S(t),I(t),R(t))\) space. Figure 3's left plot does
just that, presenting the trajectories of the simulated epidemics in
this three-dimensional space. For state space models like in our
example, given the constraint that \(S(t) + I(t)+R(t)\) is always equal
to \(N\) (the total population size) we can visual these point in a a
two-dimensional \textit{ternary} plot, as seen in Figure
\ref{fig:different-scales-tern}'s right plot. In this both of Figure
\ref{fig:different-scales-tern}'s plots we observe that the two
epidemics are on the same trajectory in this three-dimensional space. In
this simple generative example setting, being on the same trajectory
indicates that \textcolor{violet}{the} two epidemics have the same
\(R_0\) value. This can be proven mathematically.

@@

\sout{When the goal is to predict the future impact / trajectory of a epidemic,}
In addition to being used to better understand completed epidemics,
time-invariant tools can be used to aid prediction of future epidemics
by providing a state-space regions in which we expect the true epidemic
to traverse. In settings where the epidemic only generates a single
outbreak, these regions can be very telling if simulation models capture
the epidemic's structure. In \pkg{EpiCompare} we create geometric
prediction regions around all but the \(\alpha\) proportion of most
extreme simulated trajectories. These geometric regions can also be used
to compared simulation models that have different time scales,
parameters and even different statistical philosophies, through set
different distances. Although visualization is the easiest when the
epidemic has three states, prediction regions can be useful to assess
and compare simulation models where the epidemics have multiple states.

>>>>

\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way - approaches that remove some or all of the
impact of start/end times and recording time scales when preforming the
analysis. In this section we highlight some weaknesses of
time-\textit{de}pendent analysis and define the mathematical
underpinning of the time-invariant approach we take. To accomplish this
goal we first demonstrate the inability of time-dependent tools to
\sout{well quantify}\textcolor{orange}{simply capture} a classic
epidemic parameter, the reproduction number \(R_0\). Then we motivate
new time-invariant approaches for more complex situations where \(R_0\)
does not capture the complexities of outbreaks. This leads to the final
subsection that defines ways to view epidemics in a time-invariant lens
and discusses natural properties of these representations that allow for
clearer ways to compare models and epidemics.

@@

Epidemiological research is
\textcolor{orange}{often interested}\footnote{\textcolor{orange}{I used often intsead of "a lot of epidemiological research" which sounds too colloquial}}
in understanding intrinsic properties of the epidemic, independent from
when the epidemic occurred or how frequently data was collected. One of
the most famous numerical summaries of an epidemic is the reproduction
number - \(R_0\). \textcolor{orange}{} This intrinsic
property\}\footnote{\textcolor{orange}{added to shorten this long sentence and connect that it is an intrinsic property}}
is a time-invariant value that is defined as the expected number of
infections caused by a single infect or who is added to a completely
susceptible population \citep{anderson1992}. \citet{Gallagher2020}
showed that the estimation of \(R_0\) can be sensitive to time-dependent
parameters like estimation of the beginning and end of an epidemic. This
time-dependent sensitivity in estimating \(R_0\) reflects a general
problem in epidemiology surrounding how to transform time-dependent
information into insights into desirable, \textcolor{orange}{intrinsic}
properties of an epidemic.

In many situations, epidemiologists want a much deeper understand of an
epidemic than just the reproduction number. One common time-dependent
tool \textcolor{orange}{used to better understand epidemics} is a series
of time series line plots that track the proportion of the population in
a given state (e.g.~infected) at a given time. Figure
\ref{fig:different-scales-standard} visualizes two different simulated
epidemics with population states (S)usceptible, (I)nfectious and
(R)ecovered. These simulations have been generated under a discrete
approximation of \citet{Kermack1927}'s SIR model.
j\textcolor{orange}{The Kermack and McKendrick} model captures the
transitions from one state to the next as a system of ordinary
differential equations, where \(N\) is the total number of individuals,
\(\beta\) is the rate of infection, and \(\gamma\) is the rate of
recovery,

@@

Even though Figure \ref{fig:different-scales-standard}'s visual analysis
might be able to present more complex properties of the epidemic, one
might wish to understand how these two simulated epidemics' \(R_0\)
values compare. Even though these two simulated epidemics appear
different in these figures (including having different infection peaks),
there is no real intuitive way to compare the epidemics' \(R_0\) values.
In fact, both of these simulated epidemics started with the same
population (1000 total people with 10 infected) and have the same
\(R_0\) value, just with slightly different infection and recovery rates
(\(\beta_1, \gamma_1 = 0.8, 0.4\), and
\(\beta_2, \gamma_2 = .64, .32\)). Even for simple generative models,
this time-dependent visualization cannot be used to compare a very
simple but powerful numerical summary, \textcolor{orange}{$R_0$}.

@@

The faceted time series plot, like that seen in Figure
\ref{fig:different-scales-standard}, not only fails to allow the
practitioner to compare simulated epidemics' \(R_0\) values but also
presented epidemics' trajectory data so that only the marginal
information for each state can be examined at a given time. A more
holistic approach to visualizing the overall path of our simulated
epidemics is to examine \textcolor{orange}{how} each epidemic traverses
the three-dimensional \((S(t),I(t),R(t))\) space. Figure 3's left plot
does just that, presenting the trajectories of the simulated epidemics
in this three-dimensional space. For state space models like in our
example, given the constraint that \(S(t) + I(t)+R(t)\) is always equal
to \(N\) (the total population size) we can visual these point in a a
two-dimensional \textit{ternary} plot, as seen in Figure
\ref{fig:different-scales-tern}'s right plot. In this both of Figure
\ref{fig:different-scales-tern}'s plots we observe that the two
epidemics are on the same trajectory in this three-dimensional space. In
this simple generative example setting,
\textcolor{orange}{being on the same trajectory} indicates that
\textcolor{violet}{the} two epidemics have the same \(R_0\) value. This
can be proven mathematically.

@@

\sout{When the goal is to predict the future impact / trajectory of a epidemic,}
\textcolor{orange}{In addition to being used to better understand completed epidemics,}
time-invariant tools can be used to aid prediction of future epidemics
by providing a state-space regions in which we expect the true epidemic
to traverse. In settings where the epidemic only generates a single
outbreak, these regions can be very telling if simulation models capture
the epidemic's structure. In \pkg{EpiCompare} we create geometric
prediction regions around all but the \(\alpha\) proportion of most
extreme simulated trajectories. These geometric regions can also be used
to compared simulation models that have different time scales,
parameters and even different statistical philosophies, through set
different distances. Although visualization is the easiest when the
epidemic has three states, prediction regions can be useful to assess
and compare simulation models where the epidemics have multiple states.
